Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2012-12-03 MIMOSA.tex
Revised for Biostatistics

%\documentclass[11pt]{article}
\documentclass[useAMS,referee,usenatbib]{biom}
\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
%\usepackage{natbib}
\newcommand{\tr}{\mbox{tr}}
%\usepackage[margin=1in]{geometry}
%\usepackage{authblk}
%\usepackage{multicol}
%\usepackage{tikz}
\usepackage{amsmath,tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

%\usepackage[round,comma]{natbib}
%\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{booktabs}
%\usepackage{multirow}
\usepackage{url}

@@

%\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}

%\author[1]{Greg Finak}
%\author[1]{Andrew McDavid}
%\author[3]{Pratip Chattopadhyay}
%\author[3]{Maria Dominguez}
%\author[1,2]{Steve De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}
%
%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

\author{Greg Finak$^{1}$, \and
Andrew McDavid$^{1}$,
Pratip Chattopadhyay$^{3}$,
Maria Dominguez$^{3}$,
Steve De Rosa$^{1,2}$,
Mario Roederer$^{3}$,
\and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

%\date{\today}
\begin{document}

\date{{\it Received XXXXX} 2012.}

\pagerange{\pageref{firstpage}-\pageref{lastpage}}
\volume{ }
\pubyear{ }
\artmonth{ }

\doi{12.1234/j.12345-1234.2012.12345.x}

\label{firstpage}
%\maketitle
\begin{abstract}
Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

\begin{keywords}
Expectations-Maximization, Markov Chain Monte Carlo, Marginal
Likelihood, Bayesian Modeling, Hierarchical Modeling, Immunology, Flow
Cytometry, Single-Cell Gene Expression
\end{keywords}
\maketitle

@@

Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e. antigen or condition) at a time vs. unstimulated.

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. Here, by univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described in what follows.

@@

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix B, and we refer to it as the one-sided model.

@@

\subsection{Parameter estimation}
\label{s:estimation}
In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p_{si}$ and $p_{ui}$ from the null and alternative likelihoods. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix A), and are given by,

 \begin{align*}
 \begin{split}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y}_i)
=&\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
\end{split}
 \end{align*}

and
\begin{align}
\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
=&\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
\label{model2:unconstrained}
\end{split}
\end{align}
Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated. All calculations for the one-sided case are described in Web Appendix B.

\noindent\textbf{EM algorithm}\\
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\begin{split}
&\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\\ &\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\end{split}
\]
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.  Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\noindent\textbf{MCMC algorithm}\\
We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
Details about the algorithms are given in Web Appendix A. We used the
method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
the number of iterations, based on a short pilot run of the
sampler. For each dataset presented here, we calculated that no more
than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

@@

\subsection{ICS}
Using the ICS data, we performed an ROC (receiver operator characteristic) analysis to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. We considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182 being treated as true positives). The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected. Our comparisons also show that ranking based on log-fold change alone is not reliable and should not be used.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations (see Web Figure A).

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
  \node [below=of B] (E) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {E} ;
    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2. ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

@@

\backmatter

\section*{Acknowledgments}
This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH), and by grants R01 EB008400 and U01 AI068635-01 to RG, grants \#OPP38744, and \#OPP1032317 from the Bill \& Melinda Gates Foundation to VISC (Vaccine Immunology Statistical Center), grant \#OPP1032325 from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery). Study HVTN065 was conducted by the HIV Vaccine Trials Network (HVTN), and supported by the National Institute of Allergy and Infectious Diseases (NIAID). The the HVTN Laboratory Program is supported by grant \#UM1AI068618. Funding was also provided by Public Health Service grant UM1 AI068618 from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program (P30 AI027757). We thank the James B. Pendleton Charitable Trust for their generous equipment donation.

@@

Web Appendices A, B, C, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:estimation},  and \ref{s:demarkercombos} are available in the attached Web-based supplementary material.\vspace*{-8pt}
% This will be replaced at submission time
%at the Biometrics website on Wiley Online Library.

\bibliographystyle{biom}
%\bibliographystyle{unsrtnat}
\bibliography{MIMOSA}
\label{lastpage}

>>>>

\documentclass[12pt,oupdraft]{biostatistics}
%\usepackage{tikz}
%\usetikzlibrary{calc}
%\usetikzlibrary{positioning}
%\usepackage{authblk}
\usepackage{amsmath}

%\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
%\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=blue} 
\usepackage{url}
\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

%\usepackage[round,comma]{natbib}
%\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{graphicx}
%\usepackage{amssymb}
%\usepackage{booktabs}
%\usepackage{multirow}
%\usepackage{url}

@@

\title{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
%\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
%
%\author[1]{Greg Finak}
%\author[1]{Andrew McDavid}
%\author[3]{Pratip Chattopadhyay}
%\author[3]{Maria Dominguez}
%\author[1,2]{Steve De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}
%
%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

 \author{GREG FINAK$^{1\ast}$, 
 ANDREW MCDAVID$^{1}$,
 PRATIP CHATTOPADHYAY$^{3}$,
 MARIA DOMINQUEZ$^{3}$,
 STEVE DE ROSA$^{1,2}$,
 MARIO ROEDERER$^{3}$,\\
 RAPHAEL GOTTARDO$^{1}$
 \\[4pt]
\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 \\[2pt]
 {gfinak@fhcrc.org}}

\date{}
\begin{document}

%\date{{\it Received XXXXX} 2012.}

\maketitle
\footnotetext{To whom correspondence should be addressed.}

\begin{abstract}
{Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
\end{abstract}

@@

%Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e. antigen or condition) at a time vs. unstimulated.

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

%\begin{table}
%\centering
%\parbox{0.8\linewidth}{
%\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
%\centering
%\begin{tabular}{rrr}
%
%  \hline
%\multicolumn{1}{l}{}&
%\multicolumn{2}{c}{Biomarker}\\
% & Negative & Positive \\
%  \hline
%Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\
%Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\
%   \hline
%\end{tabular}
%}
%\end{table}

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described below.

@@

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.

@@

\subsection{Parameter estimation}
\label{s:estimation}
In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p_{si}$ and $p_{ui}$ from the null and alternative likelihoods. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Appendix B, Supplementary Material), and are given by,

 \begin{equation*}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y}_i)
=\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot \frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
 \end{equation*}

and
\begin{equation}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
\label{model2:unconstrained}
\end{equation}
Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated. All calculations for the one-sided case are described in Appendix A of the Supplementary Material.

\noindent\textbf{EM algorithm}\\
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\]
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.  Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\noindent\textbf{MCMC algorithm}\\
We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
the number of iterations, based on a short pilot run of the
sampler. For each dataset presented here, we calculated that no more
than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

@@

\subsection{ICS}
In the ICS data we considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182). The MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected. Our comparisons also show that ranking based on log-fold change alone is not reliable and should not be used.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations (see Supplementary Figure 1).

%\begin{figure} %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics{TIKZFig1.pdf}
%%\begin{tikzpicture} [auto, node distance=0cm]
%%\node at (0,0) (A){
%%    \begin{tikzpicture}
%%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
%%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
%%    \node at (0,1) [font=\small\sffamily] {A} ;
%%    \node at (0.5,1) [font=\small\sffamily] {B} ;
%%    \end{scope}
%%\end{tikzpicture}
%% };
%% \node [below=of A] (B) {
%% \begin{tikzpicture}
%%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
%%    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
%%    \node at (0,1) [font=\small\sffamily] {C} ;
%%    \node at (0.5,1) [font=\small\sffamily] {D} ;
%%    \end{scope}
%%    \end{tikzpicture}
%% };
%%  \node [below=of B] (E) {
%% \begin{tikzpicture}
%%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
%%    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
%%    \node at (0,1) [font=\small\sffamily] {E} ;
%%    \node at (0.5,1) [font=\small\sffamily] {F} ;
%%    \end{scope}
%%    \end{tikzpicture}
%% };
%%\end{tikzpicture}
%   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2. ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
%\label{fig:HVTN065}
%\end{figure}

We verified the assumption of a common distribution for the $p_{iu}$
across subjects by examining the empirical estimates of $p_{iu}$ and the
density of the Beta distribution with parameters $\hat{\alpha}_u,
\hat{\beta}_u$ fitted to ENV-1-PTEG stimulated
CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 2).

@@

@@

Appendix A, B, and C and Supplementary Figures 1--3 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}

\section*{Acknowledgments}
This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH) [R01 EB008400, U01 AI068635-01 to RG], grants from the Bill and Melinda Gates Foundation VISC (Vaccine Immunology Statistical Center) [grant numbers OPP38744,  OPP1032317], a grant from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery) [grant number OPP1032325].  Study HVTN065 was conducted by the HIV Vaccine Trials Network (HVTN), and supported by the National Institute of Allergy and Infectious Diseases (NIAID). The the HVTN Laboratory Program is supported by the NIH [grant  number UM1AI068618]. Funding was also provided by the Public Health Service from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program [grant numbers UM1 AI068618, P30 AI027757]. We thank the James B. Pendleton Charitable Trust for their generous equipment donation.

\bibliographystyle{biorefs}
%\bibliographystyle{unsrtnat}
\bibliography{MIMOSA}
\clearpage
